---
title: "Genome Coverage and BUSCO Filtering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Genome Coverage and BUSCO Filtering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

`fungioutline` combines lineage-aware taxon lookup with genome metadata and
BUSCO completeness metrics. This supports clade-level questions such as which
genera have public genomes, which genomes meet quality thresholds, and where
sampling gaps remain.

## Read and validate genome metadata

```{r}
genomes <- fungioutline::read_genome_metadata(
    "data-raw/fgtdb_genome_metadata.xlsx"
)

genome_validation <- fungioutline::validate_genome_metadata(
    genomes,
    error = FALSE
)

genome_validation |>
    dplyr::filter(.data$status == "fail")
```

The expected metadata includes taxonomic columns such as `phylum`, `class`,
`order`, `family`, `genus`, and genome species labels, plus BUSCO percentage and count
columns such as `C`, `F`, `M`, `complete_buscos`, and `total_buscos`.

## Classify genome quality

The default quality classes are based on BUSCO complete and missing
percentages:

```{r}
genomes_quality <- genomes |>
    fungioutline::classify_genome_quality()

genomes_quality |>
    dplyr::count(.data$genome_quality)
```

Adjust thresholds when your project uses stricter or looser standards:

```{r}
genomes_quality <- genomes |>
    fungioutline::classify_genome_quality(
        high_complete = 95,
        high_missing = 5,
        medium_complete = 80,
        medium_missing = 20
    )
```

## Filter genomes by BUSCO quality

```{r}
high_quality_genomes <- genomes |>
    fungioutline::filter_genomes_by_quality(
        min_complete = 90,
        max_missing = 10,
        require_ok = TRUE
    )
```

The function returns a tibble and preserves the input columns.

## Summarize BUSCO completeness

```{r}
fungioutline::summarize_busco(genomes, by = "phylum")

fungioutline::summarize_busco(genomes, by = c("phylum", "class"))
```

## Retrieve genomes for a clade

The packaged taxon index is used by default; pass custom genome metadata when
you are working from an external workbook.

```{r}
ascomycota_genomes <- fungioutline::get_genomes(
    "Ascomycota",
    genome_metadata = genomes
)
```

The taxon can be any supported rank. If the matched rank is broader than the
genome metadata rank, descendants are used to find matching genome rows.

## Summarize genome availability

```{r}
fungioutline::summarize_genomes(
    genomes,
    taxon = c("Ascomycota", "Basidiomycota")
)
```

The summary includes genome counts, accession counts, genome species and genus counts,
BUSCO summaries, and quality counts.

## Check descendant coverage

```{r}
coverage <- fungioutline::check_genome_coverage(
    "Ascomycota",
    genome_metadata = genomes,
    target_rank = "genus"
)

coverage |>
    dplyr::filter(!.data$has_genome)
```

## Visualize coverage and BUSCO quality

```{r}
fungioutline::plot_genome_coverage(
    "Ascomycota",
    genome_metadata = genomes,
    target_rank = "genus"
)

fungioutline::plot_busco_quality(genomes)

fungioutline::plot_busco_quality(
    genomes,
    metric = "C",
    group_col = "phylum"
)

fungioutline::plot_taxonomic_heatmap(
    "Ascomycota",
    genome_metadata = genomes,
    target_rank = "genus"
)
```

Plots are returned as ggplot objects. To save one, supply `output_path`.

```{r}
fungioutline::plot_genome_coverage(
    "Ascomycota",
    genome_metadata = genomes,
    target_rank = "genus",
    output_path = "figures/ascomycota_genome_coverage.png"
)
```
