---
title: "Getting Started with fungioutline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with fungioutline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

`fungioutline` connects a manually curated fungal taxonomic outline with genome
metadata, BUSCO quality information, optional NCBI sequence availability checks,
and visualization functions. The package is designed for tidyverse-style
analysis: functions return tibbles or ggplot objects, work in pipelines, and
avoid hidden side effects unless their purpose is explicitly to write files.

## Core objects

Most workflows use two reusable objects:

- `outline`: a standardized version of the curated outline table.
- `idx`: a long-format taxon index built from accepted names and synonyms.

```{r}
outline <- fungioutline::fungi_outline
idx <- fungioutline::fungi_taxon_index
```

The current curated outline mainly reaches genus, but the index includes a
`species` column and is designed to support `species` and `species_syn` when
those columns are added later.

## Validate the outline

Validation returns a tidy report. Use `error = FALSE` during exploration so you
can inspect all checks at once.

```{r}
validation_report <- fungioutline::validate_fungi_outline(
    outline,
    error = FALSE
)

validation_report |>
    dplyr::filter(.data$status == "fail")
```

## Look up lineages

`get_lineage()` resolves accepted names first, then synonyms when
`match_synonym = TRUE`. If a taxon name maps to multiple possible lineages,
all candidates are returned unless `best_match = TRUE`.

```{r}
fungioutline::get_lineage(
    c("Ascomycota", "Fusarium", "Gibberella"),
    taxon_index = idx
)

fungioutline::get_lineage(
    "Fusarium",
    taxon_index = idx,
    best_match = TRUE
)
```

## Add lineage columns to user data

`add_lineage()` supports bare column names and string column names, so it works
naturally in tidyverse pipelines.

```{r}
sampling_plan <- tibble::tibble(
    isolate_id = c("iso_001", "iso_002", "iso_003"),
    genus = c("Fusarium", "Agaricus", "Unknown")
)

sampling_plan |>
    fungioutline::add_lineage(taxon_col = genus, taxon_index = idx) |>
    dplyr::select(
        .data$isolate_id,
        .data$genus,
        .data$phylum,
        .data$class,
        .data$order,
        .data$family
    )
```

The same call also accepts a string:

```{r}
sampling_plan |>
    fungioutline::add_lineage(taxon_col = "genus", taxon_index = idx)
```

## Count and retrieve descendants

Use `count_taxa()` for clade-level richness and `get_descendants()` when you
need the descendant table itself.

```{r}
fungioutline::count_taxa(
    "Ascomycota",
    taxon_index = idx
)

fungioutline::get_descendants(
    "Ascomycota",
    target_rank = "genus",
    taxon_index = idx
)
```

## Visualize the outline

Plot helpers return ggplot objects and save files only when `output_path` is
supplied.

```{r}
fungioutline::plot_taxon_richness(taxon_index = idx)

fungioutline::plot_taxon_richness(
    taxon = "Ascomycota",
    by_rank = "genus",
    taxon_index = idx
)
```

## Recommended first workflow

```{r}
outline <- fungioutline::fungi_outline

validation_report <- fungioutline::validate_fungi_outline(
    outline,
    error = FALSE
)

idx <- fungioutline::fungi_taxon_index

fungioutline::get_lineage(
    c("Ascomycota", "Agaricus", "Fusarium"),
    taxon_index = idx
)

fungioutline::count_taxa("Ascomycota", taxon_index = idx)
```
