---
title: "NCBI Sequence Availability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NCBI Sequence Availability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

`fungioutline` can check whether NCBI nucleotide records are available for a
taxon or clade. The functions are serial by default, cache results, and use
`Sys.getenv("NCBI_API_KEY")` when an NCBI API key is configured.

Network examples in this vignette are not executed by default.

Taxon names are resolved before querying NCBI, so synonyms can map to accepted
names when `match_synonym = TRUE`.

## Count-only searches

Use `search_ncbi_sequence_counts()` when you only need record counts.

```{r}
counts <- fungioutline::search_ncbi_sequence_counts(
    c("Ascomycota", "Fusarium", "Agaricus"),
    db = "nuccore",
    delay_sec = 0.34
)

counts |>
    dplyr::select(
        "input_taxon",
        "accepted_name",
        "count",
        "has_sequences",
        "cached"
    )
```

The default database is `nuccore`.

## Searches with record IDs

Use `search_ncbi_sequences()` when you also want IDs up to `retmax`.

```{r}
seq_results <- fungioutline::search_ncbi_sequences(
    "Fusarium",
    db = "nuccore",
    retmax = 20,
    delay_sec = 0.34
)

seq_results |>
    dplyr::select("input_taxon", "count", "ids")
```

## Caching

Results are cached by database, query, and return limit. Choose an explicit
cache directory for reproducible projects.

```{r}
counts <- fungioutline::search_ncbi_sequence_counts(
    "Fusarium",
    cache_dir = "inst/extdata/ncbi-cache",
    delay_sec = 0.34
)
```

You can also read or write one cache item directly:

```{r}
fungioutline::cache_ncbi_result(
    key = "manual_example",
    value = list(count = 1, ids = "12345"),
    cache_dir = "inst/extdata/ncbi-cache"
)

fungioutline::cache_ncbi_result(
    key = "manual_example",
    cache_dir = "inst/extdata/ncbi-cache"
)
```

## API keys and rate limits

If available, set the NCBI API key outside R or in your interactive session:

```{r}
Sys.setenv(NCBI_API_KEY = "your_api_key_here")
```

Keep `delay_sec` above the relevant NCBI rate-limit interval for your key and
workflow. The package does not parallelize NCBI requests by default.

## Visualize NCBI availability

```{r}
fungioutline::plot_ncbi_sequence_counts(counts)
```

The returned ggplot can be saved only when you request it:

```{r}
fungioutline::plot_ncbi_sequence_counts(
    counts,
    output_path = "figures/ncbi_sequence_counts.png"
)
```

## Testing without live NCBI calls

Package tests use mocked search functions instead of live NCBI requests. For
your own tests, pass a function through `search_fun` with the same arguments:

```{r}
mock_search <- function(db, query, retmax, api_key) {
    list(
        count = 2L,
        ids = c("100", "101"),
        error = NA_character_
    )
}

fungioutline::search_ncbi_sequences(
    "Fusarium",
    use_cache = FALSE,
    delay_sec = 0,
    search_fun = mock_search
)
```
