---
title: "Phylogenomic Sampling Support"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phylogenomic Sampling Support}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

`fungioutline` is useful for building transparent phylogenomic sampling plans.
The central idea is to combine curated taxonomy, lineage-aware genome coverage,
BUSCO quality filters, and optional NCBI sequence checks before choosing taxa
for downstream analysis.

## Prepare inputs

```{r}
outline <- fungioutline::fungi_outline
idx <- fungioutline::fungi_taxon_index

genomes <- fungioutline::fgtdb_genome_metadata
```

## Define a focal clade

Start with a taxon at any supported rank.

```{r}
focal_taxon <- "Ascomycota"

lineage <- fungioutline::get_lineage(
    focal_taxon,
    taxon_index = idx,
    best_match = TRUE
)

lineage
```

## Measure taxonomic richness

```{r}
richness <- fungioutline::count_taxa(
    focal_taxon,
    taxon_index = idx
)

richness
```

Use this to decide whether sampling should be balanced by genus, family, order,
or another rank.

## Check genome coverage at the sampling rank

```{r}
coverage <- fungioutline::check_genome_coverage(
    focal_taxon,
    genome_metadata = genomes,
    target_rank = "genus",
    taxon_index = idx
)

coverage |>
    dplyr::arrange(.data$has_genome, .data$taxon_name)
```

Missing descendants can become targets for new sequencing, marker-gene
screening, or NCBI sequence checks.

## Filter candidate genomes

```{r}
candidate_genomes <- fungioutline::get_genomes(
    focal_taxon,
    genome_metadata = genomes,
    taxon_index = idx
) |>
    fungioutline::filter_genomes_by_quality(
        min_complete = 90,
        max_missing = 10,
        require_ok = TRUE
    )
```

If multiple genomes represent the same genus or species, use your project
criteria to select representatives. For example:

```{r}
representatives <- candidate_genomes |>
    dplyr::group_by(.data$genus) |>
    dplyr::arrange(dplyr::desc(.data$C), .data$n_scaffolds) |>
    dplyr::slice_head(n = 1) |>
    dplyr::ungroup()
```

## Combine coverage and quality

```{r}
coverage_quality <- coverage |>
    dplyr::left_join(
        candidate_genomes |>
            dplyr::count(.data$genus, name = "n_high_quality_genomes"),
        by = c("taxon_name" = "genus")
    ) |>
    dplyr::mutate(
        n_high_quality_genomes = tidyr::replace_na(.data$n_high_quality_genomes, 0L)
    )

coverage_quality |>
    dplyr::filter(.data$n_high_quality_genomes == 0L)
```

## Add optional NCBI sequence checks

For missing genome coverage, count whether nucleotide sequences exist in NCBI.

```{r}
missing_genome_taxa <- coverage_quality |>
    dplyr::filter(.data$n_high_quality_genomes == 0L) |>
    dplyr::pull(.data$taxon_name)

ncbi_counts <- fungioutline::search_ncbi_sequence_counts(
    missing_genome_taxa,
    taxon_index = idx,
    db = "nuccore",
    cache_dir = "inst/extdata/ncbi-cache"
)
```

## Build a sampling table

```{r}
sampling_table <- coverage_quality |>
    dplyr::left_join(
        ncbi_counts |>
            dplyr::select(
                taxon_name = .data$accepted_name,
                ncbi_count = .data$count,
                ncbi_has_sequences = .data$has_sequences
            ),
        by = "taxon_name"
    ) |>
    dplyr::mutate(
        sampling_status = dplyr::case_when(
            .data$n_high_quality_genomes > 0L ~ "genome_available",
            .data$ncbi_has_sequences ~ "sequence_available",
            TRUE ~ "sampling_gap"
        )
    )

sampling_table |>
    dplyr::arrange(.data$sampling_status, .data$taxon_name)
```

This table can be written to a project-specific file if you choose:

```{r}
readr::write_csv(
    sampling_table,
    "results/ascomycota_sampling_table.csv"
)
```

## Visual checks

```{r}
fungioutline::plot_taxon_richness(
    taxon = focal_taxon,
    taxon_index = idx
)

fungioutline::plot_genome_coverage(
    focal_taxon,
    genome_metadata = genomes,
    target_rank = "genus",
    taxon_index = idx
)

fungioutline::plot_fungioutline_framework()
```

## Reproducibility notes

Record these inputs with every sampling decision:

- Outline version ID or source Excel file.
- Genome metadata file and download date.
- BUSCO thresholds.
- NCBI database, query settings, cache directory, and query date.
- Rules used to select representative genomes.
