---
title: "Updating the Outline Excel File"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Updating the Outline Excel File}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

The curated fungal outline is intentionally edited by experts outside package
code. `fungioutline` provides the reproducible update layer around that manual
curation: import the edited Excel file, validate and standardize it, compare it
with the previous version, build a fresh taxon index, and write versioned data
plus a change history.

## Manual update cycle

1. Edit `data-raw/outline_2025.10.20.xlsx` manually.
2. Run `update_outline_data()` interactively.
3. Review the returned summary and history.
4. Run documentation and tests manually.

```{r}
outline_update <- fungioutline::update_outline_data(
    excel_path = "data-raw/outline_2025.10.20.xlsx",
    output_dir = "inst/extdata",
    overwrite_current = TRUE,
    write_versioned = TRUE,
    write_history = TRUE,
    verbose = TRUE
)

outline_update$summary
outline_update$history
```

The helper script `data-raw/update_outline.R` contains the same workflow as a
template for interactive use.

## Generated files

The update workflow can write:

- `inst/extdata/outline_current.rds`
- `inst/extdata/outline_index.rds`
- `inst/extdata/outline_<version_id>.rds`
- `inst/extdata/outline_index_<version_id>.rds`
- `inst/extdata/outline_history.tsv`

The current files are convenient package data inputs. The versioned files and
history table keep the taxonomy update reproducible.

## Read and inspect history

```{r}
history <- fungioutline::read_outline_history(
    "inst/extdata/outline_history.tsv"
)

history |>
    dplyr::count(.data$version_id, .data$change_type, name = "n_changes")
```

You can also visualize the history:

```{r}
fungioutline::plot_update_history(history)
```

## Compare two versions directly

Use `compare_outline_versions()` when you already have two outline objects in
memory.

```{r}
old_outline <- readRDS("inst/extdata/outline_current.rds")
new_outline <- fungioutline::read_fungi_outline("data-raw/outline_2025.10.20.xlsx")

history <- fungioutline::compare_outline_versions(
    old_outline = old_outline,
    new_outline = new_outline,
    version_id = "outline_review"
)

history |>
    dplyr::arrange(.data$rank, .data$taxon_name, .data$field)
```

Detected change types include:

- `added_taxon`
- `removed_taxon`
- `changed_lineage`
- `changed_synonym`
- `changed_metadata`

## Write history manually

`write_outline_history()` supports TSV, CSV, and RDS output. It creates parent
directories and appends by default.

```{r}
fungioutline::write_outline_history(
    history,
    path = "inst/extdata/outline_history.tsv",
    append = TRUE
)
```

Use `append = FALSE` only when you intentionally want to replace the existing
history file.

## Development commands

The update workflow does not run package-development commands automatically.
Run these manually after reviewing the update:

```{r}
devtools::document()
devtools::test()
devtools::check()
```
